---
title: "STA5075: Practical 4"
author: "Jessica Stow (STWJES003@myuct.ac.za)"
date: "2025-02-05"
output: 
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Produce a beautiful plot in a document:  

- Gapminder: Plot life expectancy against GDP, don’t attach, colour by continent.  
- Does a log-transformation help to bring out the information more clearly?
- R Markdown document (html, word, pdf)
- Figure caption
- Axis labels and sizing. Improve visually. Legend.
- Improve size and placement.

```{r}
library(gapminder)

str(gapminder)
nlevels(gapminder$continent) # 5 levels
levels(gapminder$continent)

summary(gapminder$gdpPercap)
summary(gapminder$lifeExp)
```


```{r}
# Plot life expectancy against GDP
plot(seq(min(gapminder$gdpPercap), max(gapminder$gdpPercap), length = 1000), 
     seq(min(gapminder$lifeExp), max(gapminder$lifeExp), length = 1000),
     type = "n",
     xlab = "GDP per capita in US $",
     ylab = "Life expectancy (years)",
     main = "Relationship between life expectancy (years) and GDP per capita (US $)")

# Add the points
# AFRICA
africa <- gapminder[gapminder$continent == "Africa",]
points(africa$gdpPercap, africa$lifeExp, col = "red")

# AMERICAS
americas <- gapminder[gapminder$continent == "Americas",]
points(americas$gdpPercap, americas$lifeExp, col = "blue")

# ASIA
asia <- gapminder[gapminder$continent == "Asia",]
points(asia$gdpPercap, asia$lifeExp, col = "green")

# EUROPE
europe <- gapminder[gapminder$continent == "Europe",]
points(europe$gdpPercap, europe$lifeExp, col = "black")

# OCEANA
oceania <- gapminder[gapminder$continent == "Oceania",]
points(oceania$gdpPercap, oceania$lifeExp, col = "purple")
```

\newpage 

```{r fig.width=10, fig.height=10, fig.align='center'}
# Plot life expectancy against log of GDP

plot(seq(min(log(gapminder$gdpPercap)), max(log(gapminder$gdpPercap)), length = 1000), 
     seq(min(gapminder$lifeExp), max(gapminder$lifeExp), length = 1000),
     type = "n",
     cex.lab = 1.2, # axis label size
     cex.axis = 0.8, 
     xlab = "GDP per capita in US $ (Log transformed)",
     ylab = "Life expectancy (years)",
     main = "Relationship between life expectancy (years) and GDP per capita (US $)")

# Add the points LOG TRANSFORMED
# AFRICA
africa <- gapminder[gapminder$continent == "Africa",]
points(log(africa$gdpPercap), africa$lifeExp, col = "red")

# AMERICAS
americas <- gapminder[gapminder$continent == "Americas",]
points(log(americas$gdpPercap), americas$lifeExp, col = "blue")

# ASIA
asia <- gapminder[gapminder$continent == "Asia",]
points(log(asia$gdpPercap), asia$lifeExp, col = "green")

# EUROPE
europe <- gapminder[gapminder$continent == "Europe",]
points(log(europe$gdpPercap), europe$lifeExp, col = "black")

# OCEANA
oceania <- gapminder[gapminder$continent == "Oceania",]
points(log(oceania$gdpPercap), oceania$lifeExp, col = "purple")

# Legend
legend("bottomright", 
       legend = c("Africa", "Americas", "Asia", "Europe", "Oceania"), 
       col = c("red", "blue", "green", "black", "purple"), 
       pch = 1, 
       cex = 0.9) 
```

\newpage

- How many different countries occur in this data set?
```{r}
length(unique(gapminder$country)) # 142 countries
```

- How many African countries?
```{r}
africa <- gapminder[gapminder$continent == "Africa",]
length(unique(africa$country)) # 52 African countries
```

- Which countries have the lowest and highest life expectancy, respectively? In which years? (There are several observations/years per country).

```{r}
lowest <- min(gapminder$lifeExp)
gapminder[gapminder$lifeExp == lowest,] # Rwanda 1992, life expectancy of 23.599 years

highest <- max(gapminder$lifeExp)
gapminder[gapminder$lifeExp == highest,] # Japan 2007, life expectancy of 82.603 years
```

