---
title: "STA5075: Practical 3"
author: "Jessica Stow (STWJES003@myuct.ac.za)"
date: "2025-02-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1
## Find all rows in airquality that have missing values.

```{r}
which(is.na(airquality)) # identify which rows have missing values
```
## Question 2
## Find mean, sd, min, max for each of temperature and ozone level.

```{r}
# Temperature
mean(airquality$Temp) # mean temperature
sd(airquality$Temp) # standard deviation of temperature
min(airquality$Temp) # minimum temperature
max(airquality$Temp) # maximum temperature

# Ozone level
mean(airquality$Ozone, na.rm = TRUE) # mean Ozone level
sd(airquality$Ozone, na.rm = TRUE) # standard deviation of Ozone level
min(airquality$Ozone, na.rm = TRUE) # minimum Ozone level
max(airquality$Ozone, na.rm = TRUE) # maximum Ozone level
```
## Question 3
## The cars data (an R data set) contains two variables: speed and distance to stop. Fit a simple linear regression model to these data, i.e. find the B estimates, using the equation above, and matrix calculations in R.

```{r}
x <- cbind(1, cars$speed) # x matrix
y <- cars$dist # y matrix

solve(t(x)%*%x)%*%(t(x)%*%y) # y intercept = -17.579095, coefficient of speed = 3.932409
```

## Question 4
## Check that you get the same B estimates as when fitting the linear regression model using lm() in R.

```{r}
m1 <- lm(dist ~ speed, data = cars)
m1$coefficients
```

## Question 5
## Mean life expectancy in South Africa? 53.99317.

```{r}
library(gapminder)
sa <- gapminder[gapminder$country == "South Africa", ] # create subset of dataset for South Africa
mean(sa$lifeExp) # mean life expectancy in South Africa = 53.99317
```

