---
title: "STA5075: Practical 15"
author: "Jessica Stow (STWJES003@myuct.ac.za)"
date: "2025-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Assume these 13 observations are from a Poisson distribution, with rate parameter $\lambda$:

```{r}
counts <- c(3, 1, 1, 3, 1, 4, 3, 2, 0, 5, 0, 4, 2) 
```

The log-likelihood is:
```{r}
pois.ll <- function(lambda, y) { 
  sum(dpois(y, lambda, log = TRUE))
}
```

The negative log-likelihood is:
```{r}
# negative log likelihood
pois.negll <- function(lambda, y) { 
  -sum(dpois(y, lambda, log = TRUE))
}
```

## Question 1a
## Use optim() and nlm() (non-linear minimization) to find the MLE for $\lambda$

```{r}
# using optim()

optim <- optim(par = 5, # Initial values for the parameters to be optimized over.
               fn = pois.negll, # A function to be minimized (or maximized), here we minimise
               y = counts, 
               method = "L-BFGS-B", # use the quasi-Newton method
               lower = 0.0001,
               upper = Inf)
optim
```

```{r}
# using nlm()
nlm <- nlm(f = pois.negll, # function to be minimized
           y = counts, 
           p = 5) # Initial values for the parameters to be optimized over.
```

We get the same number of parameters (= 2.067) and the same MLE value (=23.63712) when running the optim() and nlm() functions. 

## Question 1b
## Use Newton’s method (from scratch) to find the MLE for $\lambda$ in the above problem.

```{r}
# page 119
pois.negll <- function(lambda, y) { 
  -sum(dpois(y, lambda, log = TRUE))
}

# parameters: lambda, counts, tol (stopping value for when you find best lambda - use a while loop for stopping rules)
newton <- function(lambda, counts, tol){
  ratio <- Inf # initially set ratio to a large nr
  # keep iterating until our ratio < tol
  while(abs(ratio) > tol){
    ratio <- ((length(counts)*(lambda^2))/sum(counts)) - lambda # ratio of 1st to 2nd derivatives
    lambda <- lambda - ratio # update lambda value
  }
  print(lambda) # quit the loop when the optimum value is found
}

counts <- c(3, 1, 1, 3, 1, 4, 3, 2, 0, 5, 0, 4, 2) 
newton(lambda = 3, counts = counts, tol = 0.001)
```

