---
title: "STA5075: Practical 14"
author: "Jessica Stow (STWJES003@myuct.ac.za)"
date: "2025-02-11"
output: 
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1: Some more sampling questions!
## Assume that X is an exponential random variable with rate parameter 5. Now use Monte Carlo simulation to obtain the following quantities:


```{r}
nsamp <- 1000 # number samples drawn

set.seed(123)
exp_x <- rexp(nsamp, 5) # exponential random variable with rate parameter 5

mean_exp_x <- mean(exp_x) # E[X] 
mean_exp_x_sq <- mean(exp_x^2) # E[X^2]

# variance = E[X^2] - E[X]^2
variance = mean_exp_x_sq - (mean_exp_x)^2 # 0.04031277
variance

var(exp_x) # 0.04035312
```


```{r}
# probability between 0.1 and 0.3

# solve via integration
probability <- integrate(function(x) 5 * exp(-5 * x), lower = 0.1, upper = 0.3)$value
probability # 0.3834005

# Using pexp()
probability <- pexp(0.3, rate = 5) - pexp(0.1, rate = 5) 

abs(probability) # get the absolute value, 0.3834005
```



## Part 2a: Distribution of points in a unit square
## Consider the following set of points in the unit square. Decide whether or not they are randomly distributed in the square.

```{r}
# function to compute the euclidean distance between random points:
distance <- function(){
  # Sample n points (x;y) by sampling from two uniform distributions 
  x <- runif(n = 6, min = 0, max = 1)
  y <- runif(n = 6, min = 0, max = 1)
  # calculate euclidean distance using dist function in r
  df <- cbind(x, y)
  eu.dist <- dist(df)
  # get the mean distance
}

distance()

# must be abloe to code this from scratch

dist(x.runif, y.runif) # calculate euclidean distance
```


```{r}
dist(x.runif, y.runif)/6 # Compute the average distance between these n points

nsim = 10000

dist <- function(x, y){
  sqrt(sum((x - y)^2))
}

distance <- numeric(nsim)

for(i in 1:nsim) {
  x.runif <- runif(n = 6, min = 0, max = 1)
  y.runif <- runif(n = 6, min = 0, max = 1)
  distance[i] <- dist(x.runif, y.runif)
  return(distance)
}

hist(distance,
     main = "Euclidean distance in a uniform distribution (10 000 simulations)")

mean(distance)
sd(distance)
```


```{r}
# Given points
x <- c(0.1, .4, .45, .55, .7, .78)
y <- c(.53, .6, .7, 0.03, .48, .4)

given.dist <- dist(x, y) # calculate euclidean distance

# calculate the probability of getting this distance 
pnorm(given.dist, mean = mean(distance), sd = sd(distance)) # 0.3370951

plot(x, y, 
     xlim = c(0,1), 
     ylim = c(0,1), 
     pch = 19)
```

## Part 2b: Distribution of points on a straight line
## Are these points random between 0 and 1?

```{r}
rand_x1 <- c(.01, .02, .05, .095, 1)
xr <- seq(0, 1, by = .01)

plot(xr, rep(0, 101), 
     col = "white",
     xlab = "x", 
     ylab = "", 
     ylim = c(-.1, .1)) 

abline(h = 0)
points(rand_x1, rep(0,5), 
       col = "red", 
       pch = 19)
```

