---
title: "STA5075: Practical 12"
author: "Jessica Stow (STWJES003@myuct.ac.za)"
date: "2025-02-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


A slot machine has a single reel - think of this as a circular wheel that can turn - on which one of five different symbols can appear. These symbols are hearts, diamonds, spades, clubs and a joker.

People bet R2 at a time in order to play the game and are paid out according to the symbol that appears after spinning the machine. The reel has 50 tiles and the number of tiles that occurs on the reel for the different types is [given in the prac instructions].

## Part 1
## What is probability of getting the different winning symbols?

### 1.1 Spin the reel once and note the symbol

```{r}
# Make the reel
reel <- c("hearts", "diamonds", "spades", "clubs", "joker", "ca") # the different categories 
probs <- c(7,5,3,6,9,20)/50 # Probabilities

# Spin the wheel once and note the symbol
set.seed(1234)
spin <- sample(reel, size = 1, prob = probs)
print(spin)
```


```{r}
#probabilities of different symbols on reel
probs <- c(7,5,3,6,9,20)/50 
winnings <- c(2, 2.5, 8, 4, 2.5, 0)
spin <- sample(reel, size=1, prob = probs) 

payout <- winnings[spin]
payout
```


### 1.3 Repeat 1.1 and 1.2 above many times (10000).

```{r}
spin.10000 <- sample(reel, 
               size = 10000, 
               prob = probs, 
               replace = TRUE) 
```

### 1.4 Take note of how many times each of the symbols appear. 

```{r}
count <- as.data.frame(table(spin.10000))
print(count)
```

### 1.5 Divide by 10000. This is an estimate of the probability.

```{r}
count$Probability <- count$Freq/10000
print(count)
```

\newpage

## Part 2
## Calculate the average pay out if you spin the reel once. - i.e. one ‘play’ of the game is to spin the reel once, and you play the game 10000 times.

### 2.1 Spin the reel once.

```{r}
tile <- c("hearts", "diamonds", "spades", "clubs", "joker", "ca")

# replace reel symbols with numbers (easier subsetting)
reel <- 1:6 # let 1 be hearts, 2: diamonds, 3: spades, 4: clubs, 5: joker, 6: ca
winnings <- c(2, 2.5, 8, 4, 2.5, 0)
spin <- sample(reel, size=1, prob = probs)
```

### 2.2 Note the symbol.

```{r}
print(tile[spin])
```

### 2.3 Note the payout – store it.

```{r}
payout <- winnings[spin]
print(payout)
```

### 2.4 Repeat 2.1, 2.2 and 2.3 above many times (10000).

```{r}
set.seed(123)

spin <- sample(reel, 
               size = 10000, 
               prob = probs,
               replace = TRUE) 

payout <- winnings[spin]
```

### 2.5 You should now have 10 000 payout values.

```{r}
length(payout) # 10 000 payout values
```

### 2.6 Calculate the mean of these values. This is an estimate of the average payout if you play the game once. 
### Compare it to the true value!

The estimated true mean payout value (R1.94) differs from the actual mean payout value (in this case 1.93915). This difference is due to the fact that the outcome of the spins are independent and random, and so the expected payout fluctuates. The more the game is played, the less the outcomes fluctuate since they become more predictable as they approach the true mean payout value. This aligns with the Law of Large Numbers, which states that as more trials are conducted, the observed mean approaches the theoretical mean.

```{r}
# mean payout
mean(payout) # 1.93915

# true value
sum(winnings*probs) # 1.94
```

## Part 3
## Calculate the average total pay out if you spin the reel 10/100/1000 times - i.e. one ‘play’ of the game is to spin the reel 10/100/1000 times, and you play the game 10000 times.

```{r}
total_payout <- function(spins) {
    spin <- sample(reel, size = spins, prob = probs, replace = TRUE) 
    payout <- winnings[spin]
    cum_winnings <- sum(payout)
    return(cum_winnings)
}

set.seed(123)

# 10 Plays
plays.10 <- replicate(10000, total_payout(spins = 10))
mean(plays.10)

# 100 Plays
plays.100 <- replicate(10000, total_payout(spins = 100))
mean(plays.100)

# 1000 Plays
plays.1000 <- replicate(10000, total_payout(spins = 1000))
mean(plays.1000)
```


### 3.1 Graphically display the average total pay out after 10/100/1000 plays of the game by means of histograms and box and whisker plots.

#### 10 spins

```{r}
# Histogram
hist(plays.10,
     main = "Distribution of average total payouts for 10 spins (10 000 simulations)",
     xlab = "Total payout (Rands)",
     ylab = "Count")

# Boxplot
boxplot(plays.10,
        main = "Distribution of average total payouts for 10 spins (10 000 simulations)",
        ylab = "Total payout (Rands)")
```

#### 100 spins

```{r}
# Histogram
hist(plays.100,
     main = "Distribution of average total payouts for 100 spins (10 000 simulations)",
     xlab = "Total payout (Rands)",
     ylab = "Count")

# Boxplot
boxplot(plays.100,
        main = "Distribution of average total payouts for 100 spins (10 000 simulations)",
        ylab = "Total payout (Rands)")
```

#### 1000 spins

```{r}
# Histogram
hist(plays.1000,
     main = "Distribution of average total payouts for 1000 spins (10 000 simulations)",
     xlab = "Total payout (Rands)",
     ylab = "Count")

# Boxplot
boxplot(plays.1000,
        main = "Distribution of average total payouts for 1000 spins (10 000 simulations)",
        ylab = "Total payout (Rands)")
```


## 3.2 Describe the distribution of the average total pay out pay out for each case by calculating relevant summary statistics.

It goes without saying that as the number of plays (10/100/1000) increases, the average payout per game increases.

### 10 Plays 

```{r}
summary(plays.10)
```

The distribution is positively right-skewed which suggests that in only a small number of cases will the player finish the game with a large payout, with most games resulting in a low payout. 

### 100 Plays 

When we increase the number of spins the range of the payout widens, but the distribution remains right-skewed.

```{r}
summary(plays.100)
```

### 1000 Plays 

When we further increase the number of spins the range of the payout widens, and the distribution appears to be more normally distributed. 

```{r}
summary(plays.1000)
```

If we further increase the number of spins per play we would expect the following:  
- The average payout will increase.  
- The range of the payout values widen.  
- The spread of the values will approach a normal distribution.  
