---
title: "STA5075: Practical 1"
author: "Jessica Stow (STWJES003@myuct.ac.za)"
date: "2025-02-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1
## Calculate log(i + 1) for i = 0 to 100. The average of these values should = 3.647074.

```{r}
i <- 0:100 # create vector i

# create function
x.log <- function(x) {
    return(log(i+1))
}

x.log(i)

mean(x.log(i), na.rm = TRUE) # average =  3.647074
```

## Question 2
## Generate 10000 random values.

```{r}
set.seed(20230130) # set starting point for random number generator 
## generate 10000 values from an exponential distribution, mean = 1
y <- rexp(10000)
```

## Question 3
## Find the largest number and its position. (Answer: 9.131717, 9293)

```{r}
max(y) # find the maximum number in vector y # 9.131717
which(y == max(y)) # find its position # 9293
```
## Question 4
## How many values are > 2 (absolute and %)? (Answer: 1333, 13.33%)

```{r}
length(which(y > 2)) # 1333
length(which(y > 2))/length(y)*100 # 13.33
```

## Question 5
## y2: select every 2nd element of y, starting from 1st.

```{r}
y2 <- y[seq(from = 1, to = length(y), by = 2)]
```

## Question 6
## y3: replace values > 3 in y2 with 3. 

```{r}
y3 <- y2
y3[y3 > 3] <- 3 # replace values greater than 3 with 3
mean(y3) # average = 0.9362115
```

## Question 7
## Find the variance of y3 using vector operations. Check with var().

```{r}
sum((y3-mean(y3))^2)/(length(y3)-1) # manual calculation of the variance

var(y3) == sum((y3-mean(y3))^2)/(length(y3)-1) # TRUE
```

