---
title: "STA5075: Practical 11"
author: "Jessica Stow (STWJES003@myuct.ac.za)"
date: "2025-02-07"
output: 
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1
## Generate a vector u with 20 random values from a uniform distribution. 

```{r}
u <- runif(n = 20)
```

## Now create a new categorical variable u.cat with values depending on the values in u: ‘low’ if ui ≤ 0.3, ‘medium’ if 0.3 < ui ≤ 0.6 and ‘high’ if ui > 0.6.

```{r}
u.cat <- ifelse(u <= 0.3, "low",
                ifelse(u <= 0.6 & u > 0.3, "medium", 
                       "high"))

data.frame(u, u.cat)
```

## Question 2 
## How can you vectorize (avoid for loops) the earlier exponential maximum example?

```{r}
# EARLIER EXPONENTIAL MAXIMUM EXAMPLE

nsim <- 10000 # number of simulations
emax <- numeric(nsim) # empty vector to store values generated in for loop

for(i in 1:nsim) {
  vals <- rexp(10, rate = 5) 
  emax[i] <- max(vals)
}

# my solution - vectorize

vals <- rexp(100000, rate = 5) # generate 10 000 values 
vals <- matrix(vals, nrow = 10000, ncol = 10) # turn into a matrix with 10000 rows and 10 columns (so we can get the emax per row)

emax <- apply(vals, 1, function(x) max(x)) # get the max
print(emax)
```

## Question 3
## Create a list object with 4 matrices:

```{r}
first = matrix(38:67, 3)
second = matrix(56:91, 3)
third = matrix(82:144, 3)
fourth = matrix(46:95, 5)
listobj = list(first, second, third, fourth)

print(listobj)
```

## Extract the second column from each of the matrices.

```{r}
# use lapply() to apply function to every element of a vector or list
lapply(listobj, function(x) return(x[,2])) # extract 2nd column from each of the matrices
```

## Question 4
## Create an array (10x10x10). Fill it up with random numbers. Calculate the trace (i.e. the sum of the diagonal elements) of every matrix (10 matrices each in dimensions 1, 2 and 3).

```{r}
# create 10x10x10 array with random numbers
array <- array(rnorm(1000),
               dim = c(10, 10, 10)) 

trace <- apply(array, 3, function(x) sum(diag(x))) # for each matrix - calculate the trace (sum of diagonal elements)

trace # contains trace values for each of the 10 slices
```

