---
title: "STA5075: Practical 7"
author: "Jessica Stow (STWJES003@myuct.ac.za)"
date: "2025-02-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read the following data into R. Each time check that R has done the right thing. Leave the data as original as possible.
1. counts.xlsx (on Vula, Resources → data)
2. Tortoise data.xls, sheet ‘Tortoise measurements’ (on Vula)
3. Globular clusters: http://www.physics.mcmaster.ca/~harris/GCS_table. txt
4. Large-scale climatic index (MEI): https://www.esrl.noaa.gov/psd/enso/ mei/data/meiv2.data
5. Voice Data from Singing the Vowel ‘ooh’: http://www.statsci.org/data/ general/ooh.txt
6. Air quality data: air2.dat (on Vula)
7. 2D Pollen counts: 8.Red-cross-feb-aug-89-starts 20 Feb-1989.xls (on Vula)
8. wader counts.xls (on Vula)

```{r}
# Set working directory
getwd()

# Clear workspace
rm(list = ls())

# Check workspace is clear
ls()

library(openxlsx) # for reading .xlsx files
library(readxl) # tidyverse - works both on .xls and .xlsx files

# can set stringsAsFactors = TRUE when reading in file
```

## Question 1: counts.xlsx

```{r}
df1 <- read.xlsx("counts.xlsx", sheet = 1)

# rename columns
colnames(df1)[1] <- "observation"
colnames(df1)[4] <- "comments"

# format as date
df1$day.month.year <- as.Date(df1$day.month.year, format = "%d %b %Y")

str(df1)
View(df1)
```

## Question 2: Tortoise data.xls, sheet ‘Tortoise measurements’

```{r}
df2 <- read_excel("Tortoise data.xls", sheet = "Tortoise measurements")
str(df2)

# format as factors
df2$Sex <- as.factor(df2$Sex)
df2$Habitat <- as.factor(df2$Habitat)
df2$Behaviour <- as.factor(df2$Behaviour)

# rename last column
colnames(df2)[12] <- "Comments"

str(df2)
View(df2)
```

## Question 3: Globular clusters http://www.physics.mcmaster.ca/~harris/GCS_table.txt

```{r}
url_gcs <- "http://www.physics.mcmaster.ca/~harris/GCS_table.txt"

df3 <- read.table(url_gcs, sep = "", header = F, skip =39, nrows = 422)
head(df3)
```

## Question 4: Large-scale climatic index (MEI): https://www.esrl.noaa.gov/psd/enso/mei/data/meiv2.data
Data set unavailable

```{r}
# Data set no longer available
# url_mei <- "https://www.esrl.noaa.gov/psd/enso/mei/data/meiv2.data"

# df4 <- read.table(url_mei, sep = " ", header = FALSE, skip =1, fill = FALSE)
# View(df4)
```

## Question 5: Voice Data from Singing the Vowel ‘ooh’ http://www.statsci.org/data/general/ooh.txt

```{r}
url_ooh <- "http://www.statsci.org/data/general/ooh.txt"

df5 <- read.table(url_ooh, header = TRUE)
str(df5)
View(df5)
```

## Question 6: Air quality data: air2.dat

```{r}
df6 <- read.table("air2.dat", sep = ";")
str(df6)
```

## Question 7: 2D Pollen counts: 8.Red-cross-feb-aug-89-starts 20 Feb-1989.xls

```{r}
# 7. 2D Pollen counts: 8.Red-cross-feb-aug-89-starts 20 Feb-1989.xls (on Vula)
df7 <- read_excel("8.Red-cross-feb-aug-89-starts 20 Feb-1989.xls", 
                  skip = 2, sheet = "2 D Pollen Counts")
df7$`Site: Observatory` <- as.factor(df7$`Site: Observatory`)
df7$Category <- as.factor(df7$Category)
str(df7)
```


## Question 8: wader counts.xls

```{r}
# 8. wader counts.xls (on Vula)

df8 <- read_excel("wader counts.XLS")
str(df8)

# format as factors
df8$Site <- as.factor(df8$Site)

str(df8)
View(df8)
```
