```{r}
library(knitr)
#notice how to make a table in Rmarkdown
tile <- c("hearts", "diamonds", "spades", "clubs", "joker", "ca")
n_tiles <- c(7, 5, 3, 6, 9, 20)
win <- c(2, 2.5, 8, 4, 2.5, 0)
reel <- cbind(tile, n_tiles, win)
colnames(reel) <- c("Symbol","Number of tiles", "Payout")
kbl(reel, align = c("l","c", "c")) %>% kable_styling(bootstrap_options = 'striped')

```

```{r}
# spin 
spin <- sample(reel, size=1)
spin
```

```{r}
reel <- 1:6 #the different categories 
probs<-c(7,5,3,6,9,20 )/50
spin <- sample(reel, size = 1, prob = probs)
```

```{r}
# what does the player win

if (spin==1){payout<-2 
}else
if (spin==2) {payout<-2.5
}else
if (spin==3) {payout<-8 
}else
if (spin==4) {payout<-4 
}else
if (spin==5) {payout<-2.5 
}else
{ payout<-0 
}
```

```{r}
#probabilities of different symbols on reel
probs<-c(7,5,3,6,9,20)/50 
winnings <- c(2, 2.5, 8, 4, 2.5, 0)
spin <- sample(reel, size=1, prob = probs) 

payout <- winnings[spin]
payout
```


What is probability of getting the different winning symbols?
1.1 Spin the reel once.

```{r}
# Spin once
spin.1 <- sample(reel, size=1, prob = probs) 
```

1.2 Note the symbol.

```{r}
print(spin.1)
```

1.3 Repeat 1 and 2 above many times (10000).

```{r}
spin.10000 <- sample(reel, 
               size = 10000, 
               prob = probs, 
               replace = TRUE) 
```

1.4 Take note of how many times each of the symbols appear. 

```{r}
count <- as.data.frame(table(spin.10000))
print(count)
```

1.5 Divide by 10000. This is an estimate of the probability.

```{r}
count$Probability <- count$Freq/10000
print(count)
```
Part 2
Calculate the average pay out if you spin the reel once. - i.e. one ‘play’ of the game is to spin the reel once, and you play the game 10000 times.

2.1 Spin the reel once.

```{r}
winnings <- c(2, 2.5, 8, 4, 2.5, 0)
spin <- sample(reel, size=1, prob = probs) 

payout <- winnings[spin]
```

2.2 Note the symbol.
2.3 Note the payout – store it.
2.4 Repeat 1, 2 and 3 above many times (10000).
```{r}
winnings <- c(2, 2.5, 8, 4, 2.5, 0)
spin <- sample(reel, 
               size=10000, 
               prob = probs,
               replace = TRUE) 

payout <- winnings[spin]
```

2.5 You should now have 10 000 payout values.
```{r}
length(payout)
```

2.6 Calculate the mean of these values. This is an estimate of the average. payout if you play the game once. - Compare it to the true value!

```{r}
# mean payout
mean(payout) # 1.9685

# true value
true_mean_payout <- sum(winnings*probs) # 1.94
```

Part 3
Calculate the average total pay out if you spin the reel 10/100/1000 times - i.e. one ‘play’ of the game is to spin the reel 10/100/1000 times, and you play the game 10000 times.

```{r}
spins <- c(10, 100, 1000)
plays <- 10000

avg_payout <- function(spins) {
    spin <- sample(reel, size = spins, prob = probs, replace = TRUE) 
    payout <- winnings[spin]
    cum_winnings <- sum(winnings)
    return(cum_winnings)
}


avg_payout(spins = 10)
avg_payout(spins = 100)
avg_payout(spins = 1000)
```


3.1 Graphically display the average total pay out after 10/100/1000 plays of the game by means of histograms and box and whisker plots.

```{r}

```

3.2 Describe the distribution of the average total pay out pay out for each case by calculating relevant summary statistics.
Submit all parts in one Rmarkdown file, along with the html output of the markdown file.